Studying measurement invariance in factor analysis models via stochastic processes

### 2010-11-19 MIpaper/par_instable.tex
Modified example in manuscript, modified bib file
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@15 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

Within the general framework outlined above, model estimation
often proceeds by assuming that $\bm{\theta}_i = \bm{\theta}\
\forall\ i$, then choosing $\hat{\bm{\theta}}$ to satisfy the
criterion:
\begin{equation}
    \label{eq:score}
  \sum_{i=1}^{n} \psi(\bm{x}_i, \hat{\bm{\theta}})] = 0,    
\end{equation}
 where $\psi(\bm{x}_i, \bm{\theta})$ 
is a score function based on the model.  
For example, in factor analysis models estimated via Maximum
Likelihood, the score function is the first derivative of $\log
\text{L}(\bm{x}_i, \bm{\theta})$ with respect to $\bm{\theta}$.
General M-estimation methods may also be handled within this
framework, though we do not discuss them here.

@@

@@

\subsection{Results}
In the results section, we first describe overall results and then 
describe estimation of $\eta$ and 
isolation of model parameters violating
measurement invariance.

\subsubsection{Overall Results}
Results related to the cumulative score process are displayed in
Figure~\ref{fig:cusumex}.  In this figure, the processes
are displayed individually 
for each factor loading, with student number (ordered by age and
rescaled to (0,1)) on the x-axis and
absolute values of the cumulative scores on the y-axis.  The dashed horizontal
lines represent the critical value for rejecting $H_0$ at $\alpha=.05$
based on the double-max statistic; if any of the three processes
crosses the horizontal line, then we reject $H_0$.  It is seen that
the processes for both the second and third factor loadings
($\lambda_{21}$ and $\lambda_{31}$) cross the dashed line, indeed
implying rejection of $H_0$.  The maximum across all three processes,
corresponding to the double-max test statistic, is 1.72.  This maximum
occurs in the third graph and is represented as a solid circle, 
achieving a $p$-value of .016.

\subsubsection{Estimation of $\eta$ and Parameter Isolation}
The results of the test imply that the 
verbal scales lack measurement invariance, and the age at which the
maximum is achieved provides the estimate $\hat{\eta}$.
The student corresponding to the maximum
absolute cumulative sum (the solid circle in the third graph) is 15.8
years of age, implying one subgroup of 
students below 15.8 years and one above 15.8 years.  This agrees well
with the true $\eta$ of 16.0 years.

@@

\begin{figure}
\caption{Cumulative score processes for the verbal factor loadings,
  based on the example involving measurement invariance with respect
  to student age.  The dashed lines represent critical values for the
  test of measurement invariance at $\alpha=.05$.}
\label{fig:cusumex}
\includegraphics[height=5in]{example.pdf}
\end{figure}

>>>>

Within the general framework outlined above, model estimation
often proceeds by assuming that $\bm{\theta}_i = \bm{\theta}\
\forall\ i$, then choosing $\hat{\bm{\theta}}$ to satisfy the
criterion:
\begin{equation}
    \label{eq:score}
  \sum_{i=1}^{n} \psi(\bm{x}_i, \hat{\bm{\theta}}) = 0,    
\end{equation}
 where $\psi(\bm{x}_i, \bm{\theta})$ 
is a score function based on the model.  
For example, in factor analysis models estimated via Maximum
Likelihood, the score function is the first derivative of $\log
\text{L}(\bm{x}_i, \bm{\theta})$ with respect to $\bm{\theta}$.
General M-estimation methods may also be handled within this
framework, though we do not discuss them here.

@@

These attributes give the proposed tests important advantages over
existing tests, as existing measurement invariance methods cannot
accomplish either of the tasks outlined above.
In particular, Millsap \citeyear{Mil05} cites
``locating the invariance violation'' as a major outstanding problem
in the field.

@@

\subsection{Results}
In the results section, we first describe overall results and then 
describe estimation of $\eta$ and 
isolation of model parameters violating
measurement invariance.\footnote{R code for obtaining these results is
  available at \url{http://semtools.r-forge.r-project.org}.}

\subsubsection{Overall Results}
Results related to the cumulative score process are displayed in
Figure~\ref{fig:cusumex}.  In this figure, each panel represents the
process for a single factor loading, with student age on the x-axis and
cumulative scores on the y-axis.  The dashed horizontal
lines represent the critical value for rejecting $H_0$ at $\alpha=.05$
based on the double-max statistic; if any of the three processes
crosses the horizontal line, then we reject $H_0$.  It is seen that
the processes for both the second and third factor loadings
($\lambda_{21}$ and $\lambda_{31}$) cross the dashed line, indeed
implying rejection of $H_0$.  The absolute maximum across all three processes,
corresponding to the double-max test statistic, is 1.77.  This
statistic is obtained from the minimum of the third
panel (whose absolute value is the maximum), close to an
age of 16.  It achieves a $p$-value of .012.

\subsubsection{Estimation of $\eta$ and Parameter Isolation}
The results of the test imply that the 
verbal scales lack measurement invariance, and the age at which the
maximum is achieved provides the estimate $\hat{\eta}$.
The student corresponding to the maximum
absolute cumulative sum (the solid circle in the third graph) is 15.7
years of age, implying one subgroup of 
students below 15.7 years and one above 15.7 years.  This agrees well
with the true $\eta$ of 16.0 years.

@@

% NEED TO MODIFY
\begin{figure}
\caption{Cumulative score processes for the verbal factor loadings,
  based on the example involving measurement invariance with respect
  to student age.  Each panel represents the cumulative score process
  for a single factor loading; the red line represents the critical values for 
  the absolute maximum at $\alpha=.05$.}
\label{fig:cusumex}
\includegraphics[height=5in]{example.pdf}
\end{figure}
